---
title: "Exercises Day 3"
subtitle: "Please do not circulate."
author: 
  - "Ronald Geskus"
  - "Anh Phan Truong Quynh"
date: 'last-modified'
date-format: '[Last version:] MMMM DD, YYYY'
format: 
  html: 
    navbar: false
    df-print: paged
    code-copy: true
toc: TRUE
lof: FALSE
lot: FALSE
number-sections: FALSE
execute:
  echo: true
  eval: true
---

```{css}
/*| echo: false */
h2 {
  position: sticky;
  top: 0;
  z-index: 10;
  background-color: white;
  padding: 0.25em 0;
}
```

## Overview

Throughout these exercises, we will recreate some of the plots from [Day 2 exercises](https://oucru-modelling.github.io/R-training-2026/slides/IntroductionRDay2.html#/other-plots) using the `ggplot2` package

## Dataset

We will use the `Titanic3.xlsx` dataset

::: {.content-visible when-format="html"}
If you haven't had the data already, click the button below to download it

```{r}
#| echo: false
#| warning: false
library(downloadthis)
library(readxl)
titanic <- read_excel("data/raw/Titanic3.xlsx", na = "NA")

download_this(titanic, output_name = "Titanic3", output_extension = ".xlsx")
```
:::

We can then load the data by running

```{r}
#| message: false
library(readxl)
library(ggplot2)

# make sure to modify the file path to match where it is on your device
titanic <- read_excel("data/raw/Titanic3.xlsx", na = "NA")

head(titanic, n = 10)
```

------------------------------------------------------------------------

## Exercise 1

In this exercise, we will draw a plot exploring the relationship between `age`, `sex`, passenger class (`pclass`) and the `fare` paid for the trip.

The end result will look something like this.

```{r}
#| warning: false
#| echo: false

library(oucru)

ggplot(titanic, aes(x = age, y = fare+3, color = pclass)) +
  geom_point(size = 1) +
  geom_smooth(method = "lm") +
  geom_rug() +
  scale_y_log10("fare paid (GBP)") +
  scale_color_oucru_d() +
  facet_wrap(~sex) +
  theme_bw()
```

Seems complicated? Don't panic! the following sections will guide you through the process step-by-step.

**Step 1:** Plot the price that was paid (`fare`) as a function of `age`. Label the axis for `fare` as "fare paid (GBP)"

<details>

<summary>Expected output</summary>

```{r}
#| warning: false
#| echo: false
ggplot(titanic) +
  geom_point(aes(x = age, y = fare)) +
  ylab("fare paid (GBP)")
```

</details>

::: {.callout-tip collapse="true"}
## Hint

-   use `geom_point()` for scatter plot

-   label for y-axis and x-axis can be configure via `ylab()` and `xlab()`
:::

```{r}
#| warning: false
#| code-fold: true
#| eval: false
#| code-summary: "Reveal answer"
ggplot(titanic) +
  geom_point(aes(x = age, y = fare)) +
  ylab("fare paid (GBP)")
```

You are expected to encounter the following Warning message. Why does this occur?

```{r}
#| warning: true
#| echo: false
#| results: false
#| fig-show: hide
ggplot(titanic) +
  geom_point(aes(x = age, y = fare)) +
  ylab("fare paid (GBP)")
```

***Bonus task:*** Try adding `coord_polar()` to see what happen

```{r}
#| warning: false
#| code-fold: true
#| code-summary: "Reveal code"
#| eval: false
ggplot(titanic) +
  geom_point(aes(x = age, y = fare)) +
  ylab("fare paid (GBP)") +
  coord_polar()
```

**Step 2:** Return to the cartesian coordinate system. Add a second layer that fits the linear regression line with 95% confidence interval.

<details>

<summary>Expected output</summary>

```{r}
#| warning: false
#| echo: false
ggplot(titanic, aes(x = age, y = fare)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ylab("fare paid (GBP)")
```

</details>

::: {.callout-tip collapse="true"}
## Hint

This is considered as smoothing, use `geom_smooth(method="lm")` to fit a linear regression model
:::

```{r}
#| warning: false
#| code-fold: true
#| eval: false
#| code-summary: "Reveal answer"
ggplot(titanic, aes(x = age, y = fare)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ylab("fare paid (GBP)")
```

**Step 3:** The distribution of fare is very skewed. Therefore, it may be better to plot fare on a logarithmic scale so let's do exactly that.

<details>

<summary>Expected output</summary>

```{r}
#| warning: false
#| echo: false
ggplot(titanic, aes(x = age, y = fare+3)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_y_log10("fare paid (GBP)")
```

</details>

::: {.callout-tip collapse="true"}
## Hint

Replace `ylab()` with `scale_y_log10()`
:::

You may encounter some new Warning messages. Why do they occur, and how can we resolve them?

```{r}
#| warning: true
#| message: false
#| echo: false
#| fig-show: hide

ggplot(titanic, aes(x = age, y = fare)) +
    geom_point() +
    geom_smooth(method = "lm", formula = y ~ x) +
    scale_y_log10("fare paid (GBP)")
```

```{r}
#| warning: false
#| eval: false
#| code-fold: true
#| code-summary: "Reveal answer"
# avoid log10(0)=-Inf by adding a small value to fare
ggplot(titanic, aes(x = age, y = fare+3)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_y_log10("fare paid (GBP)")
```

**Step 4:** Now, we want to color the data points by passenger class (`pclass`), and split the calculation and plot for the regression line by these 3 groups.

<details>

<summary>Expected output</summary>

```{r}
#| warning: false
#| echo: false
ggplot(titanic, aes(x = age, y = fare+3, color = pclass)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_y_log10("fare paid (GBP)")
```

</details>

::: {.callout-tip collapse="true"}
Add `color` to the aesthetic attributes (via `aes()`)
:::

```{r}
#| warning: false
#| eval: false
#| code-fold: true
#| code-summary: "Reveal answer"
ggplot(titanic, aes(x = age, y = fare+3, color = pclass)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_y_log10("fare paid (GBP)")
```

**Step 5:** Next, we want to split the plot into separate panes for each `sex`, add a rug plot to both sides as well.

<details>

<summary>Expected output</summary>

```{r}
#| echo: false
#| warning: false
ggplot(titanic, aes(x = age, y = fare+3, color = pclass)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_rug() +
  scale_y_log10("fare paid (GBP)") +
  facet_wrap(~sex)
```

</details>

::: {.callout-tip collapse="true"}
-   Separate plots into panes via `facet_wrap()`

-   Add rug plot via `geom_rug()`
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Reveal answer"

ggplot(titanic, aes(x = age, y = fare+3, color = pclass)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_rug() +
  scale_y_log10("fare paid (GBP)") +
  facet_wrap(~sex)
```

**Step 6:** Make the plot prettier to your heart's content!

<details>

<summary>Example output</summary>

```{r}
#| echo: false
#| warning: false
ggplot(titanic, aes(x = age, y = fare+3, color = pclass)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_rug() +
  scale_y_log10("fare paid (GBP)") +
  facet_wrap(~sex) +
  scale_color_oucru_d() +
  theme_bw()
```

</details>

::: {.callout-tip collapse="true"}
## Hint

You can consider:

-   Changing color scale (e.g., `scale_color_viridis_d()`, `scale_color_manual()`)

-   Changing plot theme (e.g., `theme_bw()`, `theme_minimal()`)
:::

For the color palette, we use a custom OUCRU palette from the `oucru` package, which can be installed by running

```{r}
#| eval: false
renv::install("OUCRU-Modelling/oucru")
```

And loaded into R via

```{r}
#| output: false
library(oucru)

# The palette can be accessed via
scale_color_oucru_d()
```

You may also want to check some sample usages of the package [here](https://github.com/OUCRU-Modelling/oucru#example)

```{r}
#| eval: false
#| warning: false
#| code-fold: true
#| code-summary: "Code for example output"

ggplot(titanic, aes(x = age, y = fare, color = pclass)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_rug() +
  scale_y_log10("fare paid (GBP)") +
  facet_wrap(~sex) +
  scale_color_oucru_d() +
  theme_bw()
```

**Step 7:** Save your plot as a .png

```{r}
#| eval: false
#| warning: false
#| code-fold: true
#| code-summary: "Reveal answer"
ggsave("output/titanic_fare_plot.png", width = 7, heigh = 5)
```

------------------------------------------------------------------------

## Exercise 2

For this exercise, we will create plots to explore the `age` distribution for each `pclass`

### **Exercise 2.1. Histogram**

We can start by plotting histograms of age for each of the `pclass`.

<details>

<summary>Expected output</summary>

```{r}
#| warning: false
#| echo: false
ggplot(titanic, aes(x = age)) +
  geom_histogram() +
  facet_wrap(~pclass)
```

</details>

::: {.callout-tip collapse="true"}
## Hint

-   Use `geom_histogram()`

-   Split the plot by `pclass` via `facet_wrap()`
:::

```{r}
#| warning: false
#| eval: false
#| code-fold: true
#| code-summary: "Reveal answer"
ggplot(titanic, aes(x = age)) +
  geom_histogram() +
  facet_wrap(~pclass)
```

### **Exercise 2.2. Boxplot**

Another option is to use the boxplot. Start by making a boxplot of age (along y-axis) for each of the passenger classes (along x-axis).

```{r}
#| warning: false
#| eval: false
#| code-fold: true
#| code-summary: "Reveal answer"
ggplot(titanic) +
  geom_boxplot(
    aes(y = age, x = pclass)
  )
```

Then, add the individual data points to the boxplot. Give the data points a red color and make them (slightly) transparent.

<details>

<summary>Expected output</summary>

```{r}
#| warning: false
#| echo: false
ggplot(titanic, aes(y = age, x = pclass)) +
  geom_boxplot() +
  geom_jitter(color = "red", alpha = .2)
```

</details>

::: {.callout-tip collapse="true"}
## Hint

-   Use `geom_jitter()` instead of `geom_point()` to make it easier to discriminate between the data points.

-   Transparency level can be configured via the `alpha` aesthetic
:::

```{r}
#| warning: false
#| eval: false
#| code-fold: true
#| code-summary: "Reveal answer"
ggplot(titanic, aes(y = age, x = pclass)) +
  geom_boxplot() +
  geom_jitter(color = "red", alpha = .2)
```

### **Exercise 2.3. Violin plot**

Instead of using boxplot, you may also consider using the violin plot.

Try recreating the following plot

```{r}
#| warning: false
#| echo: false
ggplot(titanic, aes(y = age, x = pclass)) +
  geom_violin() +
  geom_jitter(color = "red", alpha = .2) +
  theme_bw()
```

```{r}
#| warning: false
#| eval: false
#| code-fold: true
#| code-summary: "Reveal answer"
ggplot(titanic, aes(y = age, x = pclass)) +
  geom_violin() +
  geom_jitter(color = "red", alpha = .2) +
  theme_bw()
```

***Bonus task:*** Customize the plot to your liking!

------------------------------------------------------------------------

## Exercise 3

Make a new graph in which we perform an exploratory analysis of the relationship between `survival` and `age`, `sex` and passenger class (`pclass`).

The end result should look something like this.

```{r}
#| echo: false
#| warning: false
ggplot(titanic, aes(x = age, y = survived, color = sex)) +
  geom_smooth(method = "loess") +
  geom_jitter(width = 0, height = .05, alpha = .5) +
  facet_wrap(~pclass) +
  coord_cartesian(ylim = c(-.1, 1.1)) +
  scale_color_oucru_d(palette = "group4") +
  theme_bw() +
  theme(legend.position = 'top')
```

A few details

-   Jitter is configured with `height=0.05` and `width=0` (experiment with different `width` values to see why you should set `width=0`)

<!-- -->

-   The smoothed layer is generated with `loess` method

-   The color is configured with oucru palette "group4"

::: {.callout-tip collapse="true"}
## Hint

List of geoms used include:

-   `geom_smooth()`

-   `geom_jitter()`

Position of legend can be configured via `theme(legend.position = ...)`

Limit for y-axis can be configured via `coord_cartesian(ylim=...)`

The oucru palette can be configured via `scale_color_oucru_d(palette = ...)`
:::

```{r}
#| eval: false
#| warning: false
#| code-fold: true
#| code-summary: "Reveal answer"
ggplot(titanic, aes(x = age, y = survived, color = sex)) +
  geom_smooth(method = "loess") +
  geom_jitter(width = 0, height = .05, alpha = .5) +
  facet_wrap(~pclass) +
  coord_cartesian(ylim = c(-.1, 1.1)) +
  scale_color_oucru_d(palette = "group4") +
  theme_bw() +
  theme(legend.position = 'top')
```


--------------------------
## R code

You can download the `.R` code file containing the answers to today's exercises using the button below

{{< downloadthis exercises_Day3_answer.R dname="exercises_Day3_answer" label="Exercise answers script" type="primary">}}
