---
title: "Exercises Day 1"
author:
- name: Ronald Geskus
  affiliations:
    - OUCRU biostat
- name: Tran Thai Hung
  affiliations:
    - OUCRU biostat
date: 'last-modified'
date-format: '[Last version:] MMMM DD, YYYY'
format: 
  pdf:
    fontsize: 12pt
toc: TRUE
lof: FALSE
lot: FALSE
number-sections: FALSE
abstract: You already made exercises 1 and 2 in the slides. For exercises 3 to 7, you will do them in RStudio. Please make sure that you have R and RStudio installed on your laptop.
execute:
  echo: true
  eval: true
---

## Exercise 1: selection within a vector

a.  Create a vector named `vec` that consists of the numbers from 11 to 30
b.  Select the 7th element of the vector
c.  Select all elements except the 15th. Hint: use a minus sign
d.  Select the 2nd and 5th element of the vector
e.  Select only the odd valued elements of `vec`.

```{r}
# a.
vec <- seq(11,30)
# b.
vec[7]
# c.
vec[-15]
# d.
vec[c(2,5)]
# e.
index <- seq(1,length(vec),by=2)
vec[index]
```

## Exercise 2: some calculations

a.  Use the elementary functions `/`, `-`, `^` and the function `sum` to calculate the mean $$\bar{x}=\sum_{i=1}^n x_i/n$$ and the standard deviation $$\sqrt{\sum_{i=1}^{n}(x_i-\bar{x})^2/(n-1)}$$ of the fare paid by the titanic passengers. 
b.  Verify your answer by using the built-in R functions `mean` and `sd`.

We create the data set  that we used in the web based interface.

```{r}
titanic <- data.frame(
    pclass=rep(c("1st","2nd","3rd"),c(4,2,4)),
    survived=c(1,0,0,0,1,1,0,0,0,0),
    name=c("Allison, Master. Hudson Trevor","Dulles, Mr. William Crothers","McCarthy, Mr. Timothy J","Walker, Mr. William Anderson", "Duran y More, Miss. Asuncion", "Mellinger, Miss. Madeleine Viol","Abbott, Master. Eugene Joseph", "Calic, Mr. Petar","Flynn, Mr. James", "Johnston, Miss. Catherine Helen"),
    age= c(0.9167,39,54,47,27,13,13,17,NA,NA),
	fare=c(151.55,29.7,51.8625,34.0208,13.8583,19.5,20.25,8.6625,7.75,23.4))
```

```{r}
# a.
fare <- titanic$fare
MeanFare <- sum(fare)/length(fare)
MeanFare
# standard deviation (in three baby steps)
numerator <- sum((fare-MeanFare)^2)
denominator <- (length(fare)-1)
StdFare  <- sqrt(numerator/denominator)
StdFare
# b.
mean(fare)
sd(fare)
```

## RStudio

### Some useful Keyboard Shortcuts in RStudio for Windows 

Over time it is useful to learn a couple of RStudio shortcuts that you frequently
use. You can find an overview of them under the **Help → Keyboard Shortcuts help** menu. More shortcuts and tips can be found on the [Appsilon](https://www.appsilon.com/post/rstudio-shortcuts-and-tips) website.

::: callout-note
On the mac, Ctrl → Cmd, Alt → Option, and Enter → Return.

-   **Ctrl+Enter**: Run current line
-   **Ctrl+Alt+I**: Insert new chunk
-   **Alt+-**: Insert assignment `<-`
-   **F1**: Show function help page when cursor is on function name
-   **Ctrl+Tab**: Switch between tabs in Source Editor
-   **Ctrl+1 / Ctrl+2**: Move focus to Source/Console
-   **Tab / Ctrl+Space**: Code suggestion/completion
-   **In Console**:
    -   Arrows **Up/Down** to navigate earlier commands
    -   **Ctrl+Up** pops up a listing of latest commands

:::

Create a folder on your computer for the exercises in this R course. Open RStudio and
create a new New Project in that folder. Open a new R Script file via the menu:

**File → New File → R Script…**

Write the code in that R Script file.


## Exercise 3: importing data from Excel and first check

We will work with a dataset that gives the survival status of passengers on the Titanic. We provide the dataset in **Excel** format ([Titanic3.xlsx](https://github.com/OUCRU-Modelling/R-training-2026/raw/refs/heads/main/slides/data/Titanic3.xlsx)). See [this link](https://hbiostat.org/data/repo/ctitanic3) for a description of the variables. Our data set has three additional columns: `dob` (date of birth), `family` (whether the person travelled with family) and `agecat` (age categorized in age ranges).


Download the dataset to the raw data subfolder of your project folder. Import the dataset
by clicking on **Import Dataset** from the **Environment** pane. Browse for the Excel
file. 

The first 50 rows of the data set are shown. Look at the code under **Code
Preview**. Assign the R object name `titanic` to the dataset under **Import Options**
and observe that the code under **Code Preview** has changed. 

Look at the mode of variable age. A problem with the mode for age is that missing
values are coded as a character value `"NA"` in our Excel file; it is not recognized
as an R type NA value. In the future, you may encounter other representations of
missing values, such as `missing`, `404`, `-99` or `unknown`, which should also be
treated as `NA`. Specify that NA represents missing values under the import
options. Observe how the code under *Code Preview** changed again. Copy the code from
**Code Preview**. Click on the "Import" button. Paste the code you have copied
earlier to your R Script File.

```{r}
#| echo: false
#| warning: false
library(readxl)
titanic <- read_excel("data/raw/Titanic3.xlsx", na="NA")
```


a.  Inspect the data set in spreadsheet-like format that shows up after importing (if it does not show up, you can click on the `titanic` object in the **Environment** pane). Do the variables make sense? Are there any weird values? Sort the data in ascending order of age by clicking on the **age** in the data set.

b.  Use the function `dim` to find the number of rows (passengers) and columns (variables) in the data set.

```{r}
dim(titanic)
```

**Answer**: 1309  rows and 17 colums

c.  Use the `str` function to obtain a summary of the basic characteristics per variable. Have a look at the `survived` column. Does it have the mode that you expect? (You can also use the `mode` function to check the modes.)

```{r}
str(titanic)
mode(titanic$survived)
```

`survived` is of mode numeric, while one would expect it to be a categorical variable
of mode character.

## Exercise 4: some columns summarized


a.  Issue the commands

```{r}
#| echo: true
head(titanic, 10)
tail(titanic, 10)
```

What do you observe?

**Answer**: First and last to rows are shown. Not all columns are shown. 

b.  The `readxl` package that is used to import the data stores the data as a `tibble`, which is a special type of format on top of the `data.frame` format to store data in R. Override the `tibble` format using the function `as.data.frame`.

```{r}
#| echo: true
titanic <- as.data.frame(titanic)
```

Run the `head` and `tail` functions again. What has changed?

```{r}
head(titanic, 10)
tail(titanic, 10)
```

**Answer**: All columns are shown. 

c.  Summarize the whole dataset using the `summary` function. What do you observe for each of the variables?

```{r summary}
summary(titanic)
```

Many are of class and mode character, and the `summary` function does not give any
information. The `dob` variable has negative values.

d.  Compute the 5%, 25%, 50%, 75% and 95% quantiles, the inter-quartile range and the standard deviation for age and fare. You may have to take a look at the `help` files for the functions `quantile`, `IQR` and `sd`.

```{r}
## some quantiles
quantile(titanic$age, probs=c(0.05,0.25,0.5,0.75,0.95), na.rm=TRUE)
quantile(titanic$fare, probs=c(0.05,0.25,0.5,0.75,0.95), na.rm=TRUE)
## the IQR (note that this is a single number)
IQR(titanic$age, na.rm=TRUE)
IQR(titanic$fare, na.rm=TRUE)
## the the standard deviation
sd(titanic$age, na.rm = TRUE)
sd(titanic$fare, na.rm = TRUE)
```

e.  For categorical variables, `summary` is not very informative. Use the `table` function to summarize the variables `pclass`, and `sex`. Also give a two-by-two table for sex and survival status using the same `table` function. Again, have a look at the help file if needed.

```{r}
## basic tables
table(titanic$pclass)
table(titanic$sex)
table(titanic$sex, titanic$survived)
## argument useNA has no added value because there are no missings in sex and survived
table(titanic$sex, titanic$survived, useNA = "always")
```

f.  The function `addmargins` adds the row sums and the column sums to the table of survival status by sex. The function `proportions` tabulates proportions instead of absolute numbers.

```{r}
#| echo: true
addmargins(table(titanic$sex, titanic$survived))
proportions(table(titanic$sex, titanic$survived))
```

Use the `proportions` function to compute the fraction (or percentage) of survivors by sex.

```{r}
## basic tables
proportions(table(titanic$sex, titanic$survived), margin = 1)
```

## Exercise 5: work with subsets

a.  Take a look at the name, and the home town/destination of all passengers who were older than 70 years. Use the appropriate selection functions.

```{r}
subset(titanic,age>70)[,c("name","home_dest")]
```

It is seen that there is one person from Uruguay in this group. Select the single record from that person. Did this person travel with relatives?

```{r}
subset(titanic, name=="Artagaveytia, Mr. Ramon")
```

**Answer**: the person did not travel with relatives

b.  Make a table of survivor status by sex, but only for the first class passengers. What do you observe? Compare this with the survival status for the 3rd class passengers.

```{r}
first <- xtabs(~sex+survived, data=titanic, subset=(pclass=="1st"))
first
third <- xtabs(~sex+survived, data=titanic, subset=(pclass=="3rd"))
third
```

**Answer**: females were much more likely to survive than males, especially among
those who travelled first class

## Exercise 6: make factor variables

a.  Make passenger class and sex into factor variables.

b.  Add a variable `status` to the Titanic data set, which gives the survival status of the passengers as a factor variable with labels "no" and "yes" describing whether individuals survived. Make the value "no" the first level.

```{r add-status}
titanic$status <- factor(titanic$survived, labels=c("no","yes"))
```

c.  Run the `summary` function again. What change do you observe compared to Exercise 4?

```{r}
summary(titanic)
```

**Answer**: Passenger class and sex are summarized properly, and status is summarized
as a categorical variable.

## Exercise 7: working with dates

a.  The variable `dob` gives the date of birth, using days since January 1st, 1960 (this is the time origin used by Stata). Transform this to a true date value using the function `as.Date` (specify the origin argument).

```{r}
#| echo: true
titanic$dob <- as.Date(titanic$dob, origin = "1960/1/1")
```

b.  The default display is "year/month/day". Make a table of the day of the month
that the passengers were born, using the `format` function. What do you observe?

```{r}
#| echo: true
table(format(titanic$dob, "%d"))
```

**Answer**: All have a day in the middle of the month

c.  What was the earliest date of birth among all the passengers on the boat. Does
this correspond to the age of the oldest person on the date that the ship sank (April
15th, 1912)? Use the functions `min` and `max`.

```{r}
#| echo: true
min(titanic$dob, na.rm = TRUE)
max(titanic$age, na.rm = TRUE)
min(titanic$dob, na.rm = TRUE) + max(titanic$age, na.rm = TRUE) * 365.25
```

Yes, it does correspond
